Clustering and Parameter Estimation for the Popularity Adjusted Block Model

### 2020-07-14 summary.Rmd
cleaned up some typos and such

**Remark**. 
Like the "ungeneralized" random dot product graph, the latent positions of a 
generalized random dot product graph (RDPG) are not unique. More specifically, 
if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \{R : R^\top I_{pq} R = I_{pq} \}$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$ (this is the indefinite orthogonal group, 
$\mathbb{O}(p, q)$).
Unlike in the RDPG case, transforming the latent positions by multiplication 
with $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances.

@@

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the regular random dot product 
graph, $Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different clusters.

@@

\begin{equation}
\begin{split}
  \min_{\theta, \phi, \psi, \tau} & ||\xi^{(1, 1)}||^2 + ||\xi^{(2, 2)}||^2 + ||\xi^{(1, 3)} - \xi^{(1, 4)}||^2 + ||\xi^{(2, 3)} + \xi^{(2, 4)}||^2 \\
  \text{s.t. } & \Xi = Z Q(\theta, \phi, \psi, \tau) \\
  & \xi^{(i, j)} \text{ is the } j^{th} \text{ column of } \Xi \text{ with rows from the } i^{th} \text{ cluster}
\end{split}
\end{equation}

@@

  XU.c <- Z.hat %*% R
  in.prods <- XU.c[clusters == 1, ] %*% t(XU.c[clusters == 2, , drop = FALSE])
  return(
    sum(XU.c[clusters == 1, 1] ** 2) + 
      sum(XU.c[clusters == 2, 2] ** 2) + 
      sum((XU.c[clusters == 1, 3] + XU.c[clusters == 1, 4]) ** 2) + 
      sum((XU.c[clusters == 2, 3] - XU.c[clusters == 2, 4]) ** 2)
  )
}

transform.embedding <- function(Z.hat, clusters) {
  # init.angles <- runif(4, -pi, pi)
  init.angles <- rep(0, 4)
  angles.1 <- optim(init.angles, function(x) obj.fun(x, Z.hat, clusters), 
                    method = 'SANN')$par
  angles.2 <- optim(angles.1, function(x) obj.fun(x, Z.hat, clusters),
                    method = 'Nelder-Mead',
                    control = list(abstol = 1e-10, maxit = 1e3))$par
  R <- construct.Q(angles.2)
  return(Z.hat %*% R)
}
```

@@

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to 90-degree rotations and reflection along axes.'}
XU.prime <- transform.embedding(Z, z)
pairs(XU.prime, col = z, asp = 1)
```

@@

```{r q_from_a, fig.cap = 'Pair plots of $\\hat{Z} $ where $\\hat{Z}$ is the ASE of an adjacency matrix of a PABM and $Q$ is estimated by minimizing objective function (1).'}
XU.hat <- transform.embedding(Zhat, z)
pairs(XU.hat, col = z, asp = 1)

>>>>

**Remark**. 
Like the RDPG, the latent positions of a GRDPG are not unique. More 
specifically, if $P_{ij} = x_i^\top I_{pq} x_j$, then we also have for any 
$Q \in \{R : R^\top I_{pq} R = I_{pq} \}$, 
$(Q x_i)^\top I_{pq} (Q x_j) = x_i^\top (Q^\top I_{pq} Q) x_j = 
x_i^\top I_{pq} x_j = P_{ij}$ (this is the indefinite orthogonal group, 
$\mathbb{O}(p, q)$).
Unlike in the RDPG case, transforming the latent positions by multiplication 
with $Q \in \mathbb{O}(p, q)$ does not necessarily maintain interpoint angles or 
distances.

@@

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the RDPG, 
$Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different clusters.

@@

\begin{equation}
\begin{split}
  \min_{\theta, \phi, \psi, \tau} & ||\xi^{(1, 2)}||^2 + ||\xi^{(2, 1)}||^2 + 
  ||\xi^{(1, 3)} - \xi^{(1, 4)}||^2 + ||\xi^{(2, 3)} + \xi^{(2, 4)}||^2 \\
  \text{s.t. } & \Xi = Z Q(\theta, \phi, \psi, \tau) \\
  & \xi^{(i, j)} \text{ is the } j^{th} \text{ column of } \Xi \text{ with rows from the } i^{th} \text{ cluster}
\end{split}
\end{equation}

@@

  XU.c <- Z.hat %*% R
  in.prods <- XU.c[clusters == 1, ] %*% t(XU.c[clusters == 2, , drop = FALSE])
  return(
    sum(XU.c[clusters == 1, 2] ** 2) + 
      sum(XU.c[clusters == 2, 1] ** 2) + 
      sum((XU.c[clusters == 1, 3] - XU.c[clusters == 1, 4]) ** 2) + 
      sum((XU.c[clusters == 2, 3] + XU.c[clusters == 2, 4]) ** 2)
  )
}

transform.embedding <- function(Z.hat, clusters, init.angles) {
  # init.angles <- runif(4, -pi, pi)
  # init.angles <- rep(0, 4)
  angles.1 <- optim(init.angles, function(x) obj.fun(x, Z.hat, clusters), 
                    method = 'SANN')$par
  angles.2 <- optim(angles.1, function(x) obj.fun(x, Z.hat, clusters),
                    method = 'Nelder-Mead',
                    control = list(abstol = 1e-10, maxit = 1e3))$par
  R <- construct.Q(angles.2)
  return(Z.hat %*% R)
}
```

@@

```{r q_trans, fig.cap = 'Pair plots of $ZQ$ where $Q$ was found by minimizing objective function (1). This embedding is equivalent to $XU$ up to 90-degree rotations and reflection along axes.'}
XU.prime <- transform.embedding(Z, z, runif(4, 0, 0))
pairs(XU.prime, col = z, asp = 1)
```

@@

```{r q_from_a, fig.cap = 'Pair plots of $\\hat{Z} $ where $\\hat{Z}$ is the ASE of an adjacency matrix of a PABM and $Q$ is estimated by minimizing objective function (1).'}
XU.hat <- transform.embedding(Zhat, z, runif(4, -pi, pi))
pairs(XU.hat, col = z, asp = 1)
